\section{Conclusion and future work}
\slabel{conclusion}

\begin{itemize}
    \item Characterize the entailments that can be represented as morphisms, both for ab- and sb-conditions
    \item Several equivalent representation of the same arrow-based condition as span-based one exist. Study normal forms. Spans with the same pushout are provably equivalent
\item Using cospans rather than spans to obtain ``light'' nested conditions. Possible, but first attempts suggest that spans are superior in terms of existence of a larger number of morphisms.
\item Other essentially equivalent definitions of Nested Application Condion exist. Explore whether our notion of morphism can be adapted to them, and if the transformation into span-based conditions is applicable as well.
\item By equipping conditions with suitable interfaces, study the algebraic structure of conditions, possibly identifying a suitable cartesian bicategory. This could provide the third structure (the combinatorial one) which together with the logical and the algebraic/categorical one presented in \cite{DBLP:journals/corr/abs-2404-18795} could lift to full FOL the triangular correspondence presented in \cite{DBLP:conf/csl/BonchiSS18} for the existential-conjunctive fragment.
\item In his work on Existential Graphs~\cite{roberts1973-the-existential-graphs-of-charles-s.-peirce}, Charles S. Peirce proposed a graphical representation of full FOL, equipped with some kinds of graph manipulations which represent sound deductions. The notion of graph is not formalized in a standard algebraic way, but certainly it would be interesting to try to equip them with a categorical structure as done for conditions in this paper.  
\item Other related work (actually works on Nested ACs): Extensions of Nested Application Conditions to arbitrary categories; Tableau%
%(orejas, leen)
; satisfiability%
%(konig et al.)
; Nested Conditions for Reactive Systems. 
\item A theory of provable entailment: backward-shift evidence is subsumed by forward-shift evidence
\item Canonical forms, e.g.\ the minimal (``thin'') span-based conditions, or mono up-arrows. We cannot restrict to these without losing provable entailment!
\item Constructions on span-based conditions, e.g.\ for conjunction (= product?) and disjunction (= coproduct?), negation (= pushing down one level), existential quantification (= back-shift?)
\end{itemize}
%
Acknowledgements: Filippo Bonchi, Nicolas Behr, Barbara König
